---
title: "Edu Suspension2"
author: "Skylyn Irby"
date: "7/9/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(ggplot2)
Suspension11 <- read.csv("suspension2011-12.csv", header = T)
Suspension12 <- read.csv("suspension2012-13.csv", header = T)
Suspension13 <- read.csv("suspension2013-14.csv", header = T)
Suspension14 <- read.csv("suspension2014-15.csv", header = T)
Suspension15 <- read.csv("suspension2015-16.csv", header = T)
Suspension16 <- read.csv("suspension2016-17.csv", header = T)

SuspensionCombined <- rbind(Suspension11, Suspension12, Suspension13, Suspension14, Suspension15, Suspension16)

length(unique(SuspensionCombined$schoolname))

length(unique(SuspensionCombined$districtcode))

write.csv(SuspensionCombined, file = "/Users/skylynolyviairby/Desktop/Edu Data Suspension/SuspensionCombined.csv")

AllData <- SuspensionCombined[-c(1:13),]
#extract when aggregate level is c

#Extracts the rows of GM for each county; 
Ccounty_GMcount <- filter(AllData, AllData$reportingcategory =="GM" , AllData$aggregatelevel == "C" )
Ccounty_GMcount

Ccounty_GFcount <- filter(AllData, AllData$reportingcategory =="GF" , AllData$aggregatelevel =="C")
Ccounty_GFcount

ConlyGMcounty <- Ccounty_GMcount[,c(1,4,6,10,11,12)]
#ConlyGMcounty

#ConlyGFcounty <- Ccounty_GFcount[, c(1,4,6,10,11,12)]
#ConlyGFcounty

#CdatframeTA <- data.frame(Ccounty_TAcount[c(2:59), "countyname"], ConlyTAcounty)
#CdatframeTA

CdatframeGM <- data.frame(Ccounty_GMcount[, ], ConlyGMcounty)
CdatframeGM

CdatframeGF <- data.frame(Ccounty_GFcount[, ], ConlyGFcounty)
CdatframeGF

#clean data for males
CdatframeMBar <- CdatframeGM[,c(1,2,4,7)]
colnames(CdatframeMBar) <- c("DistrictCode", "Year","County", "Total")
CdatframeMBar$Gender <- "Males"


#clean data for females
CdatframeFBar <- CdatframeGF[,c(1,2,4,7)]
colnames(CdatframeFBar) <- c("DistrictCode","Year","County", "Total")
CdatframeFBar$Gender <- "Females"

#combined dataframes
combinedFrameBar <- rbind(CdatframeMBar, CdatframeFBar)

geninfo <- aggregate(.~Year+Gender, data=combinedFrameBar, sum)

GenPlot <- ggplot(geninfo, aes(Year,Total)) + labs(title= "California Yearly Suspensions", subtitle = "By Gender") + geom_bar(aes(fill=Gender), stat="identity") + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
GenPlot


```


Ethnicities Bar Graph
```{r}
count_RFtot <- filter(AllData, AllData$reportingcategory == "RF" , AllData$aggregatelevel == "C") 

count_RAtot <- filter(AllData, AllData$reportingcategory == "RA" , AllData$aggregatelevel == "C") 
count_RHtot <- filter(AllData, AllData$reportingcategory == "RH" , AllData$aggregatelevel == "C") 
count_RPtot <- filter(AllData, AllData$reportingcategory == "RP" , AllData$aggregatelevel == "C") 
count_RBtot <- filter(AllData, AllData$reportingcategory == "RB" , AllData$aggregatelevel == "C") 
count_RItot <- filter(AllData, AllData$reportingcategory == "RI" , AllData$aggregatelevel == "C") 
count_RWtot <- filter(AllData, AllData$reportingcategory == "RW" , AllData$aggregatelevel == "C") 

count_RTtot <- filter(AllData, AllData$reportingcategory == "RT" , AllData$aggregatelevel == "C") 

count_RD11 <- filter(AllData, AllData$reportingcategory == "RD" , AllData$aggregatelevel == "C", AllData$ïacademicyear == "2011-12") 
#how to extract rows to get rows 2 through 59 and creates new data with only these rows

#Shows all the number of suspensions for each race in school districts

#Filipino
steprf <- count_RFtot[,c(1,4,6,7,10,11,12)]
steprf

#Asian
stepra <- count_RAtot[,c(1,4,6,7,10,11,12)]
stepra

#Hispanic
steprh <- count_RHtot[,c(1,4,6,7,10,11,12)]
steprh

#Pacific Islander
steprp <- count_RPtot[,c(1,4,6,7,10,11,12)]
steprp

#Black
steprb <- count_RBtot[,c(1,4,6,7,10,11,12)]
steprb

#Indian
stepri <- count_RItot[,c(1,4,6,7,10,11,12)]
stepri

#White
steprw <- count_RWtot[,c(1,4,6,7,10,11,12)]
steprw

#not reported
steprd <- count_RDtot[,c(1,4,6,7,10,11,12)]
steprd

#Mixed
steprt <- count_RTtot[,c(1,4,6,7,10,11,12)]
steprt

#we now need to figure out how to make a 59 by 2 new dataset that only contains the countycode and the total number of expulsions 
#creates a dataframe with only countynames for only gendermales and total expulsions 
rf <- data.frame(count_RFtot[, "districtcode"], steprf)
rf

ra <- data.frame(count_RAtot[, "districtcode"], stepra)
ra

rh <- data.frame(count_RHtot[, "districtcode"], steprh)
rh

ri <- data.frame(count_RItot[, "districtcode"], stepri)
ri

rp <- data.frame(count_RPtot[, "districtcode"], steprp)
rp

rw <- data.frame(count_RWtot[, "districtcode"], steprw)
rw

rb <- data.frame(count_RBtot[, "districtcode"], steprb)
rb

rt <- data.frame(count_RTtot[, "districtcode"], steprt)
rt

rd <- data.frame(count_RDtot[, "districtcode"], steprd)
rd
 
dfnew2 <- ra[,c(3,2,4,8)]
colnames(dfnew2)<- c("DistrictCode", "Year","County", "Total")
dfnew2$Ethnicity <- "Asian"

dfnew6 <- rb[,c(3,2,4,8)]
colnames(dfnew6)<- c("DistrictCode", "Year","County", "Total")
dfnew6$Ethnicity <- "Black"

dfnew4 <- rh[,c(3,2,4,8)]
colnames(dfnew4)<- c("DistrictCode", "Year","County", "Total")
dfnew4$Ethnicity <- "Hispanic or Latino"

dfnew8 <- rw[,c(3,2,4,8)]
colnames(dfnew8) <- c("DistrictCode", "Year","County", "Total")
dfnew8$Ethnicity <- "White"

dfnew5 <- ri[,c(3,2,4,8)]
colnames(dfnew5) <- c("DistrictCode", "Year","County", "Total")
dfnew5$Ethnicity <- "Other"

dfnew7 <- rp[,c(3,2,4,8)]
colnames(dfnew7) <- c("DistrictCode", "Year","County", "Total")
dfnew7$Ethnicity <- "Other"

dfnew3 <- rf[,c(3,2,4,8)]
colnames(dfnew3)<- c("DistrictCode", "Year","County", "Total")
dfnew3$Ethnicity <- "Other"

dfnew9 <- rt[,c(3,2,4,8)]
colnames(dfnew9) <- c("DistrictCode", "Year","County", "Total")
dfnew9$Ethnicity <- "Other"

dfnew0 <- rd[,c(3,2,4,8)]
colnames(dfnew0) <- c("DistrictCode", "Year","County", "Total")
dfnew0$Ethnicity <- "Other"

#binds the dataframes together 
ethnAll <- rbind(dfnew2,dfnew6,dfnew4,dfnew8,dfnew5, dfnew3, dfnew7, dfnew0, dfnew9)
ethnAll
#plot bar graphs 
library(ggplot2)

#combining several elments 
ethinfo <- aggregate(.~Year+Ethnicity, data=ethnAll, sum)

ethPlot <- ggplot(ethinfo, aes(Year,Total)) +labs(title="Yearly Suspension",subtitle = "By Ethnicity") + geom_bar(aes(fill=Ethnicity), stat="identity")  + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.45))

ethPlot

```


Cumulative Enrollment
```{r}
CE11 <- sum(as.numeric(as.character(onlyTAcount[1:58,11])))
CE12 <- sum(as.numeric(as.character(onlyTAcount[59:116,11])))
CE13 <- sum(as.numeric(as.character(onlyTAcount[117:174,11])))
CE14 <- sum(as.numeric(as.character(onlyTAcount[175:232,11])))
CE15 <- sum(as.numeric(as.character(onlyTAcount[233:290,11])))
CE16 <- sum(as.numeric(as.character(onlyTAcount[291:348,11])))
```

Total Suspension
```{r}

onlyTAcount <- (filter(AllData, AllData$reportingcategory =="TA", AllData$aggregatelevel =="C"))
#onlyTAcount <- filter(SuspensionCombined, SuspensionCombined$reportingcategory =="TA" && SuspensionCombined$aggregatelevel == "D1" && SuspensionCombined$totalsuspensions)

#ConlyTAcounty <- Ccounty_TAcount[c(2:59),c(10,11,12)]

#Total_Suspension  <- filter(AllData, AllData$reportingcategory =="TA" , AllData$aggregatelevel =="D1")
#Total_Suspension

#ConlyTAcounty

TA11 <- sum(as.numeric(as.character(onlyTAcount[1:58,12])))
TA12 <- sum(as.numeric(as.character(onlyTAcount[59:116,12])))
TA13 <- sum(as.numeric(as.character(onlyTAcount[117:174,12])))
TA14 <- sum(as.numeric(as.character(onlyTAcount[175:232,12])))
TA15 <- sum(as.numeric(as.character(onlyTAcount[233:290,12])))
TA16 <- sum(as.numeric(as.character(onlyTAcount[291:348,12])))

allyears <- c(TA11, TA12, TA13, TA14, TA15, TA16) 

Years <- c(2011:2016)
SuspTot <- data.frame(Years, allyears)
names(SuspTot) <- c("Years", "Totals")


totinfo <- aggregate(.~Years, data=SuspTot, sum)
options(scipen = 999)

TotPlot <- ggplot(totinfo, aes(Years,Totals)) + labs(title= "California Yearly Suspensions") + geom_bar(stat="identity") + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
TotPlot

```



```
Suspension Rate
```{r}

SR11 <- sum(as.numeric(as.character(onlyTAcount[1:58,16])))
SR12 <- sum(as.numeric(as.character(onlyTAcount[59:116,16])))
SR13 <- sum(as.numeric(as.character(onlyTAcount[117:174,16])))
SR14 <- sum(as.numeric(as.character(onlyTAcount[175:232,16])))
SR15 <- sum(as.numeric(as.character(onlyTAcount[233:290,16])))
SR16 <- sum(as.numeric(as.character(onlyTAcount[291:348,16])))

```
Violent Injury
```{r}
VI11 <- sum(as.numeric(as.character(onlyTAcount[1:58,17])))
VI12 <- sum(as.numeric(as.character(onlyTAcount[59:116,17])))
VI13 <- sum(as.numeric(as.character(onlyTAcount[117:174,17])))
VI14 <- sum(as.numeric(as.character(onlyTAcount[175:232,17])))
VI15 <- sum(as.numeric(as.character(onlyTAcount[233:290,17])))
VI16 <- sum(as.numeric(as.character(onlyTAcount[291:348,17])))

```

Violent Non-Injury
```{r}
VN11 <- sum(as.numeric(as.character(onlyTAcount[1:58,18])))
VN12 <- sum(as.numeric(as.character(onlyTAcount[59:116,18])))
VN13 <- sum(as.numeric(as.character(onlyTAcount[117:174,18])))
VN14 <- sum(as.numeric(as.character(onlyTAcount[175:232,18])))
VN15 <- sum(as.numeric(as.character(onlyTAcount[233:290,18])))
VN16 <- sum(as.numeric(as.character(onlyTAcount[291:348,18])))
```
Weapon Possession
```{r}
WP11 <- sum(as.numeric(as.character(onlyTAcount[1:58,19])))
WP12 <- sum(as.numeric(as.character(onlyTAcount[59:116,19])))
WP13 <- sum(as.numeric(as.character(onlyTAcount[117:174,19])))
WP14 <- sum(as.numeric(as.character(onlyTAcount[175:232,19])))
WP15 <- sum(as.numeric(as.character(onlyTAcount[233:290,19])))
WP16 <- sum(as.numeric(as.character(onlyTAcount[291:348,19])))
```

Drug Related
```{r}
DR11 <- sum(as.numeric(as.character(onlyTAcount[1:58,20])))
DR12 <- sum(as.numeric(as.character(onlyTAcount[59:116,20])))
DR13 <- sum(as.numeric(as.character(onlyTAcount[117:174,20])))
DR14 <- sum(as.numeric(as.character(onlyTAcount[175:232,20])))
DR15 <- sum(as.numeric(as.character(onlyTAcount[233:290,20])))
DR16 <- sum(as.numeric(as.character(onlyTAcount[291:348,20])))
```

Defiance Only
```{r}
DO11 <- sum(as.numeric(as.character(onlyTAcount[1:58,21])))
DO12 <- sum(as.numeric(as.character(onlyTAcount[59:116,21])))
DO13 <- sum(as.numeric(as.character(onlyTAcount[117:174,21])))
DO14 <- sum(as.numeric(as.character(onlyTAcount[175:232,21])))
DO15 <- sum(as.numeric(as.character(onlyTAcount[233:290,21])))
DO16 <- sum(as.numeric(as.character(onlyTAcount[291:348,21])))
```

```{r}

df <- data.frame(x = c(2011:2016), y = c(DR11, DR12, DR13, DR14, DR15, DR16), z = c(WP11, WP12, WP13, WP14, WP15, WP16))
df

#plots the number of drug suspensions 2011-2017 in a LINE GRAPH; suspension for drugs increased? why? easy access? change in laws?

#Drug Suspensions
ggplot(data=df, aes(x,y)) + geom_line(aes(x,y),color = "red") + labs(title = "Rate of Drug Related Suspensions") + labs(x = "Year", y= "Suspension for Drug Use")  + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
#Weapon Suspensions
ggplot(data=df, aes(x,y)) + geom_line(aes(x,z),color = "blue") + labs(title = "Rate of Weapon Related Suspensions") + labs(x = "Year", y= "Suspension for Weapons")  + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))


```
