---
title: "Odds Ratio"
author: "Nathalie Huerta"
date: "7/12/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r}
#got help from Jaron
```

#What is odds ratio?

An odds ratio (OR) is a measure of association between exposure and an outcoms. It is a ratio of odds describing an outcome occurring given a particular exposure as well as comparing the odds of the outcome occuring in the absence of a particular exposure. Odds Ratio is used in the following cases: 

1) Control Studies, where relative risks cannot be estimated (most commonly used)
2) Logistic Regression Analysis, ORs are computed during the analyses
3) Cross-Sectional Study Designs with Modifications
4) Cohort Study Designs with Modifications. 

An OR is the "odds of the outcome in one group by the the odds of outcome in the other group (analogous to relative risk)" (cite), ranging from zero to infinity. Hence, given condition A and condition B the odds ratio (OR) can be expressed as: 

$$OR = \frac{\frac{P(A)}{(1-P(A))}}{\frac{P(B)}{1-P(B)}}.    $$
Although odds ratio seems simple, they play several important roles. They are always valid measures of associations, but before computing an OR one must learn how to interpret it. (add more stuff later) 

#How to interpret odds ratio? 

An odds ratio is interpreted similar to relative risks (RR) but one should not confuse an odds ratio with RR. Based on a ratio obtained from condition A and condition B, OR can be interpreted in three manners: 

1) If the OR = 1, then exposure to condition A does not affect the odds of condition B.

2) If the OR >1, then there is higher odds of condition B happening when exposed to condition A. 

3) If the OR<1, then exposure is associated with lower odds of outcome. 

This may seem as a simple task but one must consider the significance of the odds ratio. If a positive OR is obtained, then does this mean we have a significant OR? Having a positive OR does not mean you have obtained a statistically significant result. In order to oder to obtain a statistically significant result one must consider the confidence interval (CI) and p-value. The 95% CI is used to measure the precision of the effect estimate of the OR. Confidence intervals are used to determine if the result from the OR lies in between the upper and lower confidence level. One should note, if the CI include 1, the OR is not statiscally significant. If the p-value > 0.05, then there is no statistical significance between the groups whereas if the p-value < 0.05, then it will be statistically significant. 

```{r}
#first you need to install the package fmsb
#install.packages("fmsb")

#then, install the fmsb library so you can run the odd ratios function 
library(fmsb)

#this will calculate the odds ratio. then call the function 
#oddsratio()
#sample code obtained from Rdocumentation 
#creates a 2 by 2 table like so this becomes the data to compute the OR
res <- oddsratio(5, 10, 85, 80)
#list the p.value, confidence interval, estimate?, method, and data.name
 str(res)
# prints out Odds ratio estimate and its significance probability 
 print(res)
#compute the odds ratio. not really sure what this part of the code dos 
 oddsratio(12, 5, 6, 12)
 oddsratio(12, 5, 6, 12, p.calc.by.independence=FALSE)
 DH <- sample(c("Disease", "Health"), 100, replace=TRUE)
 EN <- sample(c("Exposed", "Nonexposed"), 100, replace=TRUE)
 x <- table(EN, DH)
 oddsratio(x)


```

## Coding for actual dataset
###Code for odds ratio for gender for Expulsion Data
```{r}
#odds ratios for expulsions 
completemerge <- read.csv(file = "completeMerge.csv", header = T)
a <- sum(as.numeric(completemerge$Expulsions_Males))
a
b <- sum(as.numeric(completemerge$Expulsions_Females))
b
c <- sum(as.numeric(completemerge$en_female))
c
d <- sum(as.numeric(completemerge$en_male))
d

tot <- c-b
tot

tot2 <- d-a
tot2


expulmf <- matrix(c(339,308748,1473,323129),nrow=2,byrow=TRUE)
colnames(expulmf) <- c("Total Expelled","Total Not Expelled")
rownames(expulmf) <- c("Female","Male")
print(expulmf)

#why is this not running 
oddcalc <- oddsratio(expulmf)
oddcalc

#The resulting odds ratio of .24 means that females are less likely to get expelled than male. 
```

##Coding for Expulsion Race
```{r}
#odds ratio for race for expulsion
#black students vs white students
#sums the columns for the columns to find total expelled and total enrolled 
RB <- sum(as.numeric(completemerge$Expulsion_Black))
RW <- sum(as.numeric(completemerge$Expulsion_White))
enrb <- sum(as.numeric(completemerge$en_black))
enrw <- sum(as.numeric(completemerge$en_white))
#subtracts the total enrolled by ethnicity to find the total of not expelled to find OR
totexrb <- enrb - RB
totexrw <- enrw - RW
#comutes the odds ratio
expbw <- matrix(c(RB,totexrb,RW,totexrw),nrow=2,byrow=TRUE)
colnames(expbw) <- c("Total Expelled","Total Not Expelled")
rownames(expbw) <- c("Black","White")
print(expbw)

oddcalc2 <- oddsratio(expbw)
oddcalc2

#The odds is 7 times higher that a black student will get expelled compared to a white student.

#------------------------------------------------------------------------------------------------
#Black : Asian
RA <- sum(as.numeric(completemerge$Expulsion_Asian))
enra <- sum(as.numeric(completemerge$en_asian))
totexra <- enra - RA

expba <- matrix(c(RB,totexrb,RA,totexra),nrow=2,byrow=TRUE)
colnames(expba) <- c("Total Expelled","Total Not Expelled")
rownames(expba) <- c("Black","Asian")
print(expba)

oddcalcba <- oddsratio(expba)
oddcalcba

#The odds is 3 times higher that a black student will get expelled compared to an asian student.

#---------------------------------------------------------------------------------------------
RO <- sum(as.numeric(completemerge$Expulsion_Other))
enro <- sum(as.numeric(completemerge$en_other))
totexro <- enro - RO

expbo <- matrix(c(RB,totexrb,RO,totexro),nrow=2,byrow=TRUE)
colnames(expbo) <- c("Total Expelled","Total Not Expelled")
rownames(expbo) <- c("Black","Other")
print(expbo)

oddcalcbo <- oddsratio(expbo)
oddcalcbo

#The odds is 6 times higher that a black student will get expelled compared to an asian student.

#---------------------------------------------------------------------------------------------

#Black : Hispanic
RH <- sum(as.numeric(completemerge$Expulsion_Hispanic))
enrh <- sum(as.numeric(completemerge$en_hisp))
totexrh <- enrh - RH
expbh <- matrix(c(RB,totexrb,RH,totexrh),nrow=2,byrow=TRUE)
colnames(expbh) <- c("Total Expelled","Total Not Expelled")
rownames(expbh) <- c("Black","Hispanic")
print(expbh)

oddcalc4 <- oddsratio(expbh)
oddcalc4

#The odds are 5 times higher that a hispanic students will get expelled compared to a hispanic student.
```

```{r}
#hispanic vs white for expulsion
RH <- sum(as.numeric(completemerge$Expulsion_Hispanic))
enrh <- sum(as.numeric(completemerge$en_hisp))
totexrh <- enrh - RH

exphw <- matrix(c(RH,totexrh,RW,totexrw),nrow=2,byrow=TRUE)
colnames(exphw) <- c("Total Expelled","Total Not Expelled")
rownames(exphw) <- c("Hispanic","White")
print(exphw)

oddcalc3 <- oddsratio(exphw)
oddcalc3

#The odds are 1.35 times higher that a hispanic student will get expelled compared to a white student. 
#---------------------------------------------------------------------------------------------------
#Hispanic : Asian 
expha <- matrix(c(RH,totexrh,RA,totexra),nrow=2,byrow=TRUE)
colnames(expha) <- c("Total Expelled","Total Not Expelled")
rownames(expha) <- c("Hispanic","Asian")
print(expha)

oddcalcha <- oddsratio(expha)
oddcalcha

#The resulting odds ratio of .64 means that Hispanics are less likely to get expelled than Asian.
#look up at results for OR when it is less than 1 

#--------------------------------------------------------------------------------------------------
#Hispanic : Black for expulsion 

exphb <- matrix(c(RH,totexrh,RB,totexrb),nrow=2,byrow=TRUE)
colnames(exphb) <- c("Total Expelled","Total Not Expelled")
rownames(exphb) <- c("Hispanic","Black")
print(exphb)

oddcalchb <- oddsratio(exphb)
oddcalchb

#The resulting odds ratio of .19 means that Hispanics are less likely to get expelled than blacks.

#--------------------------------------------------------------------------------------------------
#Hispanic : Other for expulsion
expho <- matrix(c(RH,totexrh,RO,totexro),nrow=2,byrow=TRUE)
colnames(expho) <- c("Total Expelled","Total Not Expelled")
rownames(expho) <- c("Hispanic","Other")
print(expho)

oddcalcho <- oddsratio(expho)
oddcalcho

#The odds are 1.18 times higher that a hispanic student will get expelled compared to other. 

```

```{r}
#Asian : Black for Expulsion
expab <- matrix(c(RA,totexra,RB,totexrb),nrow=2,byrow=TRUE)
colnames(expab) <- c("Total Expelled","Total Not Expelled")
rownames(expab) <- c("Asian","Black")
print(expab)

oddcalcab <- oddsratio(expab)
oddcalcab

#The resulting odds ratio of .30 means that Asian students are less likely to get expelled than black students.

#----------------------------------------------------------------------------------------------------
#Asian : Hispanic for Expulsion 
expah <- matrix(c(RA,totexra,RH,totexrh),nrow=2,byrow=TRUE)
colnames(expah) <- c("Total Expelled","Total Not Expelled")
rownames(expah) <- c("Asian","Hispanic")
print(expah)

oddcalcah <- oddsratio(expah)
oddcalcah

#The odds are 1.56 times higher that a Asian student will get expelled compared to a Hispanic student. 
#---------------------------------------------------------------------------------------------------
#Asian : Other
expao <- matrix(c(RA,totexra,RO,totexro),nrow=2,byrow=TRUE)
colnames(expao) <- c("Total Expelled","Total Not Expelled")
rownames(expao) <- c("Asian","Other")
print(expao)

oddcalcao <- oddsratio(expao)
oddcalcao

#The odds are 1.84 times higher that a Asian student will get expelled compared to a student who is other.

#----------------------------------------------------------------------------------------------------
#Asian : White
expaw <- matrix(c(RA,totexra,RW,totexrw),nrow=2,byrow=TRUE)
colnames(expaw) <- c("Total Expelled","Total Not Expelled")
rownames(expaw) <- c("Asian","White")
print(expaw)

oddcalcaw <- oddsratio(expaw)
oddcalcaw

#The odds are 2 times higher that an Asian student will get expelled compared to a student who is white.

```

```{r}
#odds ratio for White race for expulsion
#White : Asian
expwa <- matrix(c(RW,totexrw,RA,totexra),nrow=2,byrow=TRUE)
colnames(expwa) <- c("Total Expelled","Total Not Expelled")
rownames(expwa) <- c("White","Asian")
print(expwa)

oddcalcwa <- oddsratio(expwa)
oddcalcwa

#The resulting odds ratio of .47 means that White students are less likely to get expelled than asian students.

#---------------------------------------------------------------------------------------------------
#Whites : Black for expulsion 
expwb <- matrix(c(RW,totexrw,RB,totexrb),nrow=2,byrow=TRUE)
colnames(expwb) <- c("Total Expelled","Total Not Expelled")
rownames(expwb) <- c("White","Black")
print(expwb)

oddcalcwb <- oddsratio(expwb)
oddcalcwb

#The resulting odds ratio of .14 means that White students are less likely to get expelled than Black students. 

#----------------------------------------------------------------------------------------------------
#Whites : Hispanic for expulsion 
expwh <- matrix(c(RW,totexrw,RH,totexrh),nrow=2,byrow=TRUE)
colnames(expwh) <- c("Total Expelled","Total Not Expelled")
rownames(expwh) <- c("White","Hispanic")
print(expwh)

oddcalcwh <- oddsratio(expwh)
oddcalcwh

#The resulting odds ratio of .74 means that White students are less likely to get expelled than Hispanic students.

#--------------------------------------------------------------------------------------------------
#White : Other
expwo <- matrix(c(RW,totexrw,RO,totexro),nrow=2,byrow=TRUE)
colnames(expwo) <- c("Total Expelled","Total Not Expelled")
rownames(expwo) <- c("White","Other")
print(expwo)

oddcalcwo <- oddsratio(expwo)
oddcalcwo

#The resulting odds ratio of .87 means that White students are less likely to get expelled than other students.

```


##Odds Ratios for Dropout Data for gender 
```{r}
#odds ratio for dropout 
#females vs males
dropfem <- sum(as.numeric(completemerge$do_female))
dropmal <- sum(as.numeric(completemerge$do_male))

droptotfem <- c - dropfem
droptotmal <- d - dropmal

dropfm <- matrix(c(dropfem, droptotfem, dropmal, droptotmal),nrow=2,byrow=TRUE)
colnames(dropfm) <- c("Total Dropout","Total Who Did Not Dropout")
rownames(dropfm) <- c("Female","Male")
print(dropfm)

oddcalcdrop <- oddsratio(dropfm)
oddcalcdrop

#The resulting odds ratio of .71 means that females were less likely to dropout than males. 
```

```{r}
#odds ratio for race/ethnicity for dropout 
#black : white for dropout
doRB <- sum(as.numeric(completemerge$do_black))
doRW <- sum(as.numeric(completemerge$do_white))
enrb <- sum(as.numeric(completemerge$en_black))
enrw <- sum(as.numeric(completemerge$en_white))

dototexrb <- enrb - doRB
dototexrw <- enrw - doRW
#comutes the odds ratio
doexpbw <- matrix(c(doRB,dototexrb,doRW,dototexrw),nrow=2,byrow=TRUE)
colnames(doexpbw) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpbw) <- c("Black","White")
print(doexpbw)

oddrop <- oddsratio(doexpbw)
oddrop

#The odds is 4 times higher that a black student will dropout when compared to a white student. 

#-------------------------------------------------------------------------------------------------

#Black : Asian 
doRA <- sum(as.numeric(completemerge$do_asian))
dototra <- enra - doRA

doexpba <- matrix(c(doRB,dototexrb,doRA,dototra),nrow=2,byrow=TRUE)
colnames(doexpba) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpba) <- c("Black","Asian")
print(doexpba)

oddropba <- oddsratio(doexpba)
oddropba

#The odds is 11 times higher that a black student will dropout when compared to an asian student.

#---------------------------------------------------------------------------------------------------

#Black : Hispanic 
doRH <- sum(as.numeric(completemerge$do_hisp))
doenrh <- sum(as.numeric(completemerge$en_hisp))
dototrh <- doenrh - doRH

doexpbh <- matrix(c(doRB,dototexrb,doRH,dototrh),nrow=2,byrow=TRUE)
colnames(doexpbh) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpbh) <- c("Black","Hispanic")
print(doexpbh)

oddropbh <- oddsratio(doexpbh)
oddropbh

#The odds is 1.84 times higher that a black student will dropout when compared to a hispanic student.

```

```{r}
#odds ratio for race/ethnicity for dropout 
#Asian : Black

doexpab <- matrix(c(doRA,dototra,doRB,dototexrb),nrow=2,byrow=TRUE)
colnames(doexpab) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpab) <- c("Asian","Black")
print(doexpab)

oddropab <- oddsratio(doexpab)
oddropab

#---------------------------------------------------------------------------------------------------

#Asian : Hispanic

doexpah <- matrix(c(doRA,dototra,doRH,dototexrh),nrow=2,byrow=TRUE)
colnames(doexpah) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpah) <- c("Asian","Hispanic")
print(doexpah)

oddropah <- oddsratio(doexpah)
oddropah

#The resulting odds ratio of .16 means that Asians were less likely to dropout than Hispanic. 

#---------------------------------------------------------------------------------------------------

#Asian : White

doexpaw <- matrix(c(doRA,dototra,doRW,dototexrw),nrow=2,byrow=TRUE)
colnames(doexpaw) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpaw) <- c("Asian","White")
print(doexpaw)

oddropaw <- oddsratio(doexpaw)
oddropaw

#The resulting odds ratio of .37 means that Asians were less likely to dropout than Whites.

```


```{r}
#hispanic: white 

doexphw <- matrix(c(doRH,dototexrh,doRW,dototexrw),nrow=2,byrow=TRUE)
colnames(doexphw) <- c("Total Dropout","Total Non- Dropout")
rownames(doexphw) <- c("Hispanic","White")
print(doexphw)

oddrophw <- oddsratio(doexphw)
oddrophw

#The odds are 2 times higher that a hispanic student will get expelled compared to a white student. 

#------------------------------------------------------------------------------------------------

#Hispanic: Black
doexphb <- matrix(c(doRH,dototexrh,doRB,dototexrb),nrow=2,byrow=TRUE)
colnames(doexphb) <- c("Total Dropout","Total Non- Dropout")
rownames(doexphb) <- c("Hispanic","Black")
print(doexphb)

dooddcalhb <- oddsratio(doexphb)
dooddcalhb


#the resulting odds ratio of .54 means that hispanic were less likely to dropout than black students.

#------------------------------------------------------------------------------------------------

#Hispanic : Asia 

doexpha <- matrix(c(doRH,dototexrh,doRA,dototexra),nrow=2,byrow=TRUE)
colnames(doexpha) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpha) <- c("Hispanic","Asia")
print(doexpha)

dooddcalha <- oddsratio(doexpha)
dooddcalha

#The odds are 6 times higher that a hispanic student will dropout compared to an asian student. 
```

```{r}
#white : Asian 

doexpwa <- matrix(c(doRW,dototexrw,doRA,dototexra),nrow=2,byrow=TRUE)
colnames(doexpwa) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpwa) <- c("White","Asia")
print(doexpwa)

dooddcalwa <- oddsratio(doexpwa)
dooddcalwa

#The odds are 3 times higher that a white student will dropout compared to an asian student. 

#------------------------------------------------------------------------------------------------

#white : Black

doexpwb <- matrix(c(doRW,dototexrw,doRB,dototexrb),nrow=2,byrow=TRUE)
colnames(doexpwb) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpwb) <- c("White","Black")
print(doexpwb)

dooddcalwb <- oddsratio(doexpwb)
dooddcalwb

#The resulting odds ratio of .24 means that whites were less likely to dropout than blacks.

#------------------------------------------------------------------------------------------------

#white hispanic 

doexpwh <- matrix(c(doRW,dototexrw,doRH,dototexrh),nrow=2,byrow=TRUE)
colnames(doexpwh) <- c("Total Dropout","Total Non- Dropout")
rownames(doexpwh) <- c("White","Hispanic")
print(doexpwh)

dooddcalwh <- oddsratio(doexpwh)
dooddcalwh

#The resulting odds ratio of .45 means that whites were less likely to dropout than hispanics.

```

